Letter: Performance analysis of \\ high-resolution ice sheet simulations

### 2022-06-26 paper/perfmod.tex
the first three pages are as solid as they are likely to be

Ice sheet models involve climate interactions (coupling), especially via surface mass balance, on time scales which are dominated by an annual cycle and certain longer scales.  We define $q$ as the number of ice-dynamical time steps per year needed to capture this coupling.  Some typical values $q=0.1 \,\text{a}^{-1}, 1 \,\text{a}^{-1}, 12 \,\text{a}^{-1}$ correspond to decadal, yearly, and monthly coupling frequency, respectively.  Note that energy balance and degree-day schemes for computing surface mass balance generally have much shorter time scales, but here $q$ only describes the frequency on which ice geometry is updated from an ice velocity solution and the latest surface mass balance values.

As already noted, current-technology ice sheet models use explicit time-stepping which is only conditionally stable \citep{LeVeque2007}.  For the spatial resolutions used in present-day scientific applications, we observe that maintenance of explicit time-stepping stability requires time steps substantially shorter than $1/q$ model years.  For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ \citep{Bueleretal2005,HindmarshPayne1996} where $D$ is a representative diffusivity value from the diffusivity formula in equation \eqref{eq:siamasscontinuity}.

For Stokes dynamics the stability of explicit time-stepping is largely unexplored in any precise sense, but we propose that an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, computed from simulated ice velocities, represents the \emph{optimistic} paradigm.  The corresponding \emph{pessimistic} paradigm enforces $\Delta t < O(D^{-1} \Delta x^2)$, using diagnostic diffusivity values computed in the SIA manner.

Explicit time-stepping with hybrid and higher-order schemes is better-studied than for Stokes dynamics, especially over horizontal resolutions relevant to whole ice sheets.  Some hybrid schemes apply the pessimistic time step as an adaptive restriction \citep{Winkelmannetal2011}.  The optimistic time step is supported for a certain higher-order DIVA scheme \citep[see equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest $\Delta t = O(\Delta x^{1.6})$ (Figure 3(a)).

Unconditionally-stable implicit schemes also have a maximum time step restriction, namely $1/q$ itself.  This restriction reflects the simulation purpose and not the maintenance of stability.  Thus, for an implicit scheme the frequency of climate coupling determines the time-stepping simulation cost, through the need to solve the coupled mass/momentum equations with frequency $q$.

Regardless of time-stepping design, for each explicit time step the velocity (or velocity/pressure) solution of the stress balance equations requires effort, namely the solution of (spatial) partial differential equations.  For the Stokes model we suppose one such solution requires $O(m^{1+\alpha})$ work, which we describe the the number floating point operations in our simple performance analysis.

Of course, the power $\alpha$ depends on the design of the stress balance solver.  A direct linear algebra Stokes solver implementation might yield $\alpha=3$, or $\alpha \approx 2$ when exploiting sparsity, but a multigrid method, such as the algebraic multigrid \citep{Trottenbergetal2001} solver by \cite{IsaacStadlerGhattas2015}, reduces $\alpha$.  Their Antarctic ice sheet results show Newton-Krylov iterations \citep{Bueler2021} growing only slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$ \citep[see Table 8.1]{IsaacStadlerGhattas2015}.
% see data/scalingisaac.m
A geometric multigrid method applied to the higher-order stress balance equations \citep{BrownSmithAhmadia2013} suggests $\alpha$ also close to zero for certain simplified geometries.\footnote{The \cite{IsaacStadlerGhattas2015} and \cite{BrownSmithAhmadia2013} results also show good parallel scaling, something not considered in our simplified analysis.}  At the even easier end, the isothermal SIA velocity computation is a trivialization of the Stokes problem in which the vertically-averaged velocity at each node is computed by a pointwise formula, $O(1)$ work; a velocity solution requires only $O(m)$ work ($\alpha=0$).

>>>>

Ice sheet models resolve climate interactions (coupling), especially via surface mass balance, on time scales which are dominated by an annual cycle and certain longer scales.  Let $q$ be the number of ice-dynamical time steps per year needed to capture this coupling.  Typical values $q=0.1 \,\text{a}^{-1}, 1 \,\text{a}^{-1}, 12 \,\text{a}^{-1}$ correspond to decadal, yearly, and monthly coupling frequency, respectively.  Note that energy balance and degree-day schemes for computing surface mass balance generally have much shorter time scales, but here $q$ only describes the frequency on which ice geometry is updated using velocity, i.e.~via solution of the mass continuity or surface kinematical equation.

As already noted, current-technology ice sheet models use explicit time-stepping which is only conditionally stable.  For the spatial resolutions used in present-day scientific applications, we observe that maintenance of explicit time-stepping stability requires steps substantially shorter than $1/q$ model years.

For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ \citep{Bueleretal2005,HindmarshPayne1996} where $D$ is a representative diffusivity value (from equation \eqref{eq:siamasscontinuity}).  For Stokes dynamics the stability of explicit time-stepping is largely unexplored in any precise sense, but we propose that an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, or computed from modeled ice velocities, represents the \emph{optimistic} paradigm.  The corresponding \emph{pessimistic} paradigm enforces $\Delta t < O(D^{-1} \Delta x^2)$, using (otherwise diagnostic) diffusivity values computed in the SIA manner.

Explicit time-stepping with hybrid and higher-order schemes is better-studied than for Stokes dynamics, especially over horizontal resolutions relevant to whole ice sheets.  Some hybrid schemes apply the pessimistic time step as an adaptive restriction \citep{Winkelmannetal2011}.  The optimistic time step is supported for a certain higher-order DIVA scheme \citep[see equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest an intermediate power $\Delta t = O(\Delta x^{1.6})$ (Figure 3(a)).

Unconditionally-stable implicit schemes also have a maximum time step restriction, namely $1/q$ itself.  This restriction reflects the simulation purpose and not the maintenance of stability.  Indeed, for an implicit scheme the frequency of climate coupling determines the time-stepping simulation cost, according to the solution cost of coupled mass and momentum equations.

For each explicit time step the computational cost of a velocity (or velocity/pressure) solution of the stress balance equations is determined by solver design.  For the Stokes model we suppose one such solution requires $O(m^{1+\alpha})$ floating point operations.\footnote{Many considerations are suppresed in our simple analysis, including memory latency, memory bandwidth, and process/thread/GPU parallelism.}

The power $\alpha$ depends on the design of the stress balance solver.  A direct linear algebra Stokes solver implementation might yield $\alpha=3$, or $\alpha \approx 2$ when exploiting sparsity, but a multigrid method, such as the algebraic multigrid \citep{Trottenbergetal2001} solver by \cite{IsaacStadlerGhattas2015}, reduces $\alpha$.  Their Antarctic ice sheet results show Newton-Krylov iterations \citep{Bueler2021} growing only slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$ \citep[see Table 8.1]{IsaacStadlerGhattas2015}.
% see data/scalingisaac.m
A geometric multigrid method applied to the higher-order stress balance equations \citep{BrownSmithAhmadia2013} suggests $\alpha$ also close to zero for certain simplified geometries.\footnote{The \cite{IsaacStadlerGhattas2015} and \cite{BrownSmithAhmadia2013} results also show good parallel scaling, something not considered in our simplified analysis.}  At the even easier end, the isothermal SIA velocity computation is a trivialization of the Stokes problem in which the vertically-averaged velocity at each node is computed by a pointwise formula, $O(1)$ work; a velocity solution requires only $O(m)$ work ($\alpha=0$).
